An Empirical Analysis of C Preprocessor Use

### 2000-01-15 macros/papers/emp-use-2.tex
symbol -> identifier in lots of places;  some places I left symbol when the context justified it.

%% mcat_SYMBOL_UNKNOWN: 1798
%%   100%  unknown_symbol (1798)
% @mcat_SYMBOL_UNKNOWN = qw( catSYMBOL_UNKNOWN );
\item[Unknown symbol]
  The macro expands to a single identifier that is not defined in the package
  or in any library header files included by the package.  The symbol may
  be defined by compiler command arguments or may be used inside a
  conditional compilation guard because it is only meaningful 
  with a particular architecture, system, or library (for which we did not
  have header files available).

  Unknown symbols can also be local variables, or variables or functions that we failed to
  parse.  Our approximate parser can succeed where an exact parse would not
  (as for non-syntactic code that deviates from the language grammar
  or for entities interrupted by preprocessor
  directives), but occasionally it fails to recognize declarations or
  definitions.

@@

  Some categorization failures resulted from limitations of our parser, which does not
  handle pasting (token concatenation) and stringization (converting a macro
  argument to a C string); together, these represent 3\% of failures.  Handling of partial
  entities is incomplete, so case labels (4\% of failures), partial expressions (6\%),
  and some declarations (14\%) are left uncategorized, as are bodies that
  pass a non-first-class object (such as a type or operator) to another
  macro (7\%).  The biggest problem is macros with adjacent symbols, which
  is generally not permitted in C\@ (35\% of macros).  These macro bodies
  often use a macro argument that is a statement or operator or expand to
  a partial declaration or cast.  Examples include
\begin{verbatim}
    #define zdbug(s1) if (zdebug) syslog s1;
    #define PNG_EXPORT(type,symbol) __declspec(dllexport) type symbol
    #define gx_device_psdf_common gx_device_vector_common; psdf_distiller_params params
\end{verbatim}
  There were also some cases in which uses of macros categorized as
  ``other'' caused macro bodies in which they appeared to be similarly
  categorized.

@@

We expected problems dealing with macros that use stringization (conversion
of a preprocessor symbol to a C string) and pasting (creating a new
identifier from two existing identifiers), the two explicit
extra-linguistic features of Cpp.  However, these features appear in only
0.07\% of all macro definitions.  Far more prevalent, and more problematic
for program understanding tools, is exploitation of Cpp's lack of structure
to effect mechanisms not available in C\@.  Cpp's inputs and outputs are
uninterpreted token streams, so Cpp can perform unstructured transformations
using non-first-class or partial syntactic constructs, such as types or
partial declarations.

@@

\begin{description}
\item[Free variables]\label{page:freevar}
  The macro body uses as a subexpression (that is, applies an operator or
  function to) a symbol that is not a formal argument, a variable defined
  in the macro body, a global variable, a reserved word, or a function, macro, or typedef
  name.  Such symbols are typically local variables at the site of
  macro invocation.  Uses of local variables (in which the local definition
  in scope at the point of use captures the free variable in the macro
  body) can produce dynamic scoping, which C does not directly support.
  Examples of this usage include
\begin{verbatim}
    #define INV_LINE(line) &invisible_line[L_OFFSET((line), wrap_offset)]
    #define atime mailfiles[i]->access_time
\end{verbatim}

@@

\item[Pasting]\label{def:pasting}
  The body uses symbol pasting ({\tt \#\#}), which treats its arguments not
  as tokens but as strings, constructing a new token out of their
  concatenation.  After {\tt \#define \verb|_SIZEOF|(x) \verb|sz_|\#\#x},
  the macro invocation {\tt \verb|_SIZEOF|(int)} expands to the 
  symbol {\tt \verb|sz_int|}.  The resulting symbol might appear literally, or
  only as a pasted symbol, at its other uses.  Pasting is often
  abstracted out into a separate macro\,---\,such as {\tt \#define
  \verb|__CONCAT|(x,y) x \#\# y}\,---\,but such pasting macros are used less
  than typical macros defined in our test suite.

@@

While ANSI~C uses {\tt \#\#} for pasting, in K\&R~C one abuts two symbols,
with an empty comment in between, so that their expansions form a new
identifier when the compiler removes the comment.  For instance, in K\&R~C,
{\tt to/**/ken} is interpreted as a single token, and macros might be
expanded on either side of the comment as well.  This construct does not
perform merging in newer implementations, so we warn users of its
appearance.  We do not report it in our statistics because use of {\tt
/**/}-style pasting is rarely a bug and is not uncommon, especially in {\tt
CONCAT} macros that provide portability across older and newer versions of
the preprocessor.

@@

The least frequently redefined macros are those categorized as unknown
symbol.  This is partially due to our method of coalescing multiple
definitions:  macro definitions categorized as unknown symbol are overridden
by definitions with any other categorization (see
Figure~\ref{fig:category-lub}).  Our
analysis included enough library header files to include some
recognizable definition of most common macros.

@@

% \begin{figure}
%   {\small\centerline{\input{tbl-summarize-frequency.tex}}}
%   
%   \captionsmall{Summary of
%     Figures~\ref{fig:freq-def-cat},~\ref{fig:freq-ddf-cat},
%     and~\ref{fig:freq-use-cat}.  The table is by macro name.  Add percentages?
%     [[The 1.0 distinct defs for unknown symbol is surprising; does it indicate
%     that it is the *same* unknown symbol in each definition?
%     But a name is only in the category if none of its defs was
%     recognized.  Double-check this!]]}
%   \label{fig:freq-sum-cat}
% \end{figure}

@@

\item If one is an unknown symbol (or the name of an undefined macro), use
  the other on the premise that the unseen definition is likely to be
  similar to the available one, which is true for well-behaved macros.

@@

\item If one is an ambiguous list of space-separated symbols and the other
  is a reserved word or type, use the latter.  Sequences of symbols are
  difficult to definitively identify in isolation, but the other
  definitions of the same name indicate the intended usage.

@@

The category breakdown by macro name (detailed in the legend of
Figure~\ref{fig:freq-use-cat}) differs from the by-definition breakdown of
Figure~\ref{fig:categorization} in several ways.  The number of null
definitions is lower, as null definitions are often found in conjunction
with other types of definition and are eliminated by the category merging.
(Macros defined only via null definitions are generally used only in Cpp
conditionals.)  The number of statements is lower, largely because
some macros names with statement definitions have other, incompatible
definitions, so the macro name is categorized as ``other''.  The percentage
of unknown symbols is higher because such macros tend to have very few
definitions, so are more prominent in a breakdown by name than by
definition.  The number of ``other'' categorizations increased because it
includes any macro with a definition categorized as ``other'' as well as any
with incompatible definitions.

@@

\item[Redefinition suppression] These guards test non-definedness of
  symbol, and control only a definition of the same symbol, thus avoiding
  preprocessor warnings about a redefinition of a name (e.g.,
  \texttt{\#ifndef FOO} followed by \texttt{\#define FOO ...} and
  \texttt{\#endif}).  
  The purpose is to provide a default value used unless another part of the
  system, or the compilation command, specifies another value.

>>>>

%% mcat_SYMBOL_UNKNOWN: 1798
%%   100%  unknown_symbol (1798)
% @mcat_SYMBOL_UNKNOWN = qw( catSYMBOL_UNKNOWN );
\item[Unknown identifier]
  The macro expands to a single identifier that is not defined in the package
  or in any library header files included by the package.  The symbol may
  be defined by compiler command arguments or may be used inside a
  conditional compilation guard because it is only meaningful 
  with a particular architecture, system, or library (for which we did not
  have header files available).

  Unknown identifiers can also be local variables, or variables or functions that we failed to
  parse.  Our approximate parser can succeed where an exact parse would not
  (as for non-syntactic code that deviates from the language grammar
  or for entities interrupted by preprocessor
  directives), but occasionally it fails to recognize declarations or
  definitions.

@@

  Some categorization failures resulted from limitations of our parser, which does not
  handle pasting (token concatenation) and stringization (converting a macro
  argument to a C string); together, these represent 3\% of failures.  Handling of partial
  entities is incomplete, so case labels (4\% of failures), partial expressions (6\%),
  and some declarations (14\%) are left uncategorized, as are bodies that
  pass a non-first-class object (such as a type or operator) to another
  macro (7\%).  The biggest problem is macros with adjacent identifiers, which
  is generally not permitted in C\@ (35\% of macros).  These macro bodies
  often use a macro argument that is a statement or operator or expand to
  a partial declaration or cast.  Examples include
\begin{verbatim}
    #define zdbug(s1) if (zdebug) syslog s1;
    #define PNG_EXPORT(type,symbol) __declspec(dllexport) type symbol
    #define gx_device_psdf_common gx_device_vector_common; psdf_distiller_params params
\end{verbatim}
  There were also some cases in which uses of macros categorized as
  ``other'' caused macro bodies in which they appeared to be similarly
  categorized.

@@

We expected problems dealing with macros that use stringization (conversion
of a preprocessor macro argument to a C string) and pasting (creating a new
identifier from two existing identifiers), the two explicit
extra-linguistic features of Cpp.  However, these features appear in only
0.07\% of all macro definitions.  Far more prevalent, and more problematic
for program understanding tools, is exploitation of Cpp's lack of structure
to effect mechanisms not available in C\@.  Cpp's inputs and outputs are
uninterpreted token streams, so Cpp can perform unstructured transformations
using non-first-class or partial syntactic constructs, such as types or
partial declarations.

@@

\begin{description}
\item[Free variables]\label{page:freevar}
  The macro body uses as a subexpression (that is, applies an operator or
  function to) an identifier that is not a formal argument, a variable defined
  in the macro body, a global variable, a reserved word, or a function, macro, or typedef
  name.  Such identifiers are typically local variables at the site of
  macro invocation.  Uses of local variables (in which the local definition
  in scope at the point of use captures the free variable in the macro
  body) can produce dynamic scoping, which C does not directly support.
  Examples of this usage include
\begin{verbatim}
    #define INV_LINE(line) &invisible_line[L_OFFSET((line), wrap_offset)]
    #define atime mailfiles[i]->access_time
\end{verbatim}

@@

\item[Pasting]\label{def:pasting}
  The body uses symbol pasting ({\tt \#\#}), which treats its arguments not
  as tokens but as strings, constructing a new token out of their
  concatenation.  After {\tt \#define \verb|_SIZEOF|(x) \verb|sz_|\#\#x},
  the macro invocation {\tt \verb|_SIZEOF|(int)} expands to the 
  identifier {\tt \verb|sz_int|}.  The resulting identifier might appear literally, or
  only as a pasted identifier, at its other uses.  Pasting is often
  abstracted out into a separate macro\,---\,such as {\tt \#define
  \verb|__CONCAT|(x,y) x \#\# y}\,---\,but such pasting macros are used less
  than typical macros defined in our test suite.

@@

While ANSI~C uses {\tt \#\#} for pasting, in K\&R~C one abuts two identifiers,
with an empty comment in between, so that their expansions form a new
identifier when the compiler removes the comment.  For instance, in K\&R~C,
{\tt to/**/ken} is interpreted as a single token, and macros might be
expanded on either side of the comment as well.  This construct does not
perform merging in newer implementations, so we warn users of its
appearance.  We do not report it in our statistics because use of {\tt
/**/}-style pasting is rarely a bug and is not uncommon, especially in {\tt
CONCAT} macros that provide portability across older and newer versions of
the preprocessor.

@@

The least frequently redefined macros are those categorized as unknown
identifier.  This is partially due to our method of coalescing multiple
definitions:  macro definitions categorized as unknown identifiers are overridden
by definitions with any other categorization (see
Figure~\ref{fig:category-lub}).  Our
analysis included enough library header files to include some
recognizable definition of most common macros.

@@

% \begin{figure}
%   {\small\centerline{\input{tbl-summarize-frequency.tex}}}
%   
%   \captionsmall{Summary of
%     Figures~\ref{fig:freq-def-cat},~\ref{fig:freq-ddf-cat},
%     and~\ref{fig:freq-use-cat}.  The table is by macro name.  Add percentages?
%     [[The 1.0 distinct defs for unknown identifier is surprising; does it indicate
%     that it is the *same* unknown identifier in each definition?
%     But a name is only in the category if none of its defs was
%     recognized.  Double-check this!]]}
%   \label{fig:freq-sum-cat}
% \end{figure}

@@

\item If one is an unknown identifier (or the name of an undefined macro), use
  the other on the premise that the unseen definition is likely to be
  similar to the available one, which is true for well-behaved macros.

@@

\item If one is an ambiguous list of space-separated identifiers and the other
  is a reserved word or type, use the latter.  Sequences of identifiers are
  difficult to definitively identify in isolation, but the other
  definitions of the same name indicate the intended usage.

@@

The category breakdown by macro name (detailed in the legend of
Figure~\ref{fig:freq-use-cat}) differs from the by-definition breakdown of
Figure~\ref{fig:categorization} in several ways.  The number of null
definitions is lower, as null definitions are often found in conjunction
with other types of definition and are eliminated by the category merging.
(Macros defined only via null definitions are generally used only in Cpp
conditionals.)  The number of statements is lower, largely because
some macros names with statement definitions have other, incompatible
definitions, so the macro name is categorized as ``other''.  The percentage
of unknown identifiers is higher because such macros tend to have very few
definitions, so are more prominent in a breakdown by name than by
definition.  The number of ``other'' categorizations increased because it
includes any macro with a definition categorized as ``other'' as well as any
with incompatible definitions.

@@

\item[Redefinition suppression] These guards test non-definedness of
  identifiers, and control only a definition of the same identifier, thus avoiding
  preprocessor warnings about a redefinition of a name (e.g.,
  \texttt{\#ifndef FOO} followed by \texttt{\#define FOO ...} and
  \texttt{\#endif}).  
  The purpose is to provide a default value used unless another part of the
  system, or the compilation command, specifies another value.
